Data Exploration: IUCN and AquaMaps

### 2015-11-18 data_explore.Rmd
updating plots and such

### theme for species range maps - based on generic plot theme
ggtheme_map <- ggtheme_plot +
  theme(axis.text = element_blank()) +       
  scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
  scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2))

@@

  spp_plot <- ggplot(spp_pts, aes(x = x, y = y)) +
    ggtheme_map + 
    geom_raster(aes(fill = presence), alpha = .5) +
    scale_fill_manual(values = c('iucn' = 'blue', 'am' = 'red', 'both' = 'mediumorchid')) +
    borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
    labs(title = spp, x = NULL, y = NULL) 
}

@@

summary(lm(total_area ~ sm_perc, spp_list))

@@

scatter_arearatio_vs_overlap <- ggplot(spp_list, aes(x = area_ratio, y = sm_perc, color = log_total_area)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline() + 
  geom_vline(x = area_ratio_quartiles,
             color = 'grey50', linetype = 'dashed') +
  geom_hline(y = pct_overlap_quartiles,
             color = 'grey50', linetype = 'dashed') +
  geom_point(data = spp_list_label, color = 'tomato2') + 
  geom_text(data = spp_list_label, 
            aes(label = sciname_label), 
            color = 'tomato2',
            fontface = 'bold.italic',
            size = 2,
            hjust = .25,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap + 
  geom_text(x = 15, y = 1, label = lm_eqn2(spp_list), parse = TRUE, color = 'tomato2')

@@

scatter_arearatio_vs_overlap2 <- ggplot(spp_list, aes(x = area_ratio, y = sm_perc, color = sm_range)) + 
  ggtheme_plot + 
  geom_point() + 
  geom_abline() + 
  geom_vline(x = area_ratio_quartiles,
             color = 'grey50', linetype = 'dashed') +
  geom_hline(y = pct_overlap_quartiles,
             color = 'grey50', linetype = 'dashed') +
  geom_point(data = spp_list_label, color = 'grey30') + 
  geom_text(data = spp_list_label, 
            aes(label = sciname_label), 
            color = 'grey30',
            fontface = 'bold.italic',
            size = 2,
            hjust = .25,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

scatter_arearatio_vs_overlap2 <- scatter_arearatio_vs_overlap2 + 
  geom_text(x = 15, y = 1, label = lm_eqn2(spp_list), parse = TRUE, color = 'grey30')

ggsave('figures/AreaRatiovsPercOverlap2.png', 
       width = 6, height = 4)

print(scatter_arearatio_vs_overlap2)

summary(lm(area_ratio ~ sm_perc, spp_list))

@@

barchart_spp_gp_quads <- ggplot(spp_gp_quadrants %>% 
                                  transform(spp_group = reorder(spp_group, pct_q1)), 
                                aes(x = spp_group, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  geom_bar(stat = 'bin') +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = .05), hjust = 0) +
  coord_flip() +
  labs(x = 'Species Group', 
       y = 'Relative number of species per quadrant', 
       title = 'Species group breakdown by quadrant')

@@

``` {r, echo = FALSE, warning = FALSE, message = FALSE}

# Regression

ggsave('figures/CorrelationAreavsPercOverlap.png')

summary(lm(total_area ~ sm_perc, spp_list))
```

@@

Percent coverage as function of area, clipping to area below 100,000,000 km^2:
``` {r, echo = FALSE, warning = FALSE, message = FALSE}
# remove all area greater than 100000000

@@

### create dataframe that gives me all species, their total number of cells, 
### total number of cells >= 0.4 prob, difference between the two and threat
spp_df3 <- spp_cells %>%
  group_by(species_id) %>%
  summarize(cells_0.4 = sum(probability >= 0.4),
            cells_t = sum(probability >= 0)) %>%
  mutate(prop = cells_0.4 / cells_t) %>%
  left_join(spp %>%
#              filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX', 'DD', 'NE')), 
              filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX')), 
            by = c('species_id' = 'SPECIESID')) %>%
  select(sciname, species_id, cells_0.4, cells_t, prop, iucn_code)

hist_area_vs_threshold <- ggplot(spp_df3, aes(prop, fill = iucn_code)) + 
  ggtheme_plot + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.05, alpha = .7) + 
  labs(x = 'Proportion of range area included when selecting a threshold of 0.4', 
       y = 'Proportion of all aquamaps species (17,348 total)')

@@

>>>>

### theme for species range maps - based on generic plot theme
ggtheme_map <- ggtheme_plot +
  theme(axis.text = element_blank())

@@

  spp_plot <- ggplot(spp_pts, aes(x = x, y = y)) +
    ggtheme_map + 
    geom_raster(aes(fill = presence), alpha = .5) +
    scale_fill_manual(values = c('iucn' = 'blue', 'am' = 'red', 'both' = 'mediumorchid')) +
    borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
    scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
    scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2)) +
    labs(title = spp, x = NULL, y = NULL) 
}

@@

# summary(lm(total_area ~ sm_perc, spp_list))

@@

scatter_arearatio_vs_overlap <- ggplot(spp_list, aes(x = area_ratio, y = sm_perc, color = log_total_area)) + 
  ggtheme_plot + 
  ### color the quadrant backgrounds:
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "green4")  + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax = 100, alpha = .3, fill= "greenyellow") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax = 100, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orange") + 
    annotate("rect", xmin = area_ratio_quartiles[2],  xmax =   0, 
                     ymin = pct_overlap_quartiles[2], ymax =   0, alpha = .3, fill= "orangered2") + 
  geom_point() + 
  geom_vline(x = area_ratio_quartiles[c(1, 3)],
             color = 'grey50', linetype = 'dashed') +
  geom_hline(y = pct_overlap_quartiles[c(1, 3)],
             color = 'grey50', linetype = 'dashed') +
  geom_point(data = spp_list_label, color = 'tomato2') + 
  geom_text(data = spp_list_label, 
            aes(label = sciname_label), 
            color = 'tomato2',
            fontface = 'bold.italic',
            size = 2,
            hjust = .25,
            vjust = 1.5) +
  labs(x = 'Range area ratio (smaller:larger) * 100%', 
       y = 'Percent overlap smaller range in larger', 
       title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')

scatter_arearatio_vs_overlap <- scatter_arearatio_vs_overlap + 
  geom_text(x = 40, y = 5, label = lm_eqn2(spp_list), parse = TRUE, color = 'tomato2')

@@

# scatter_arearatio_vs_overlap2 <- ggplot(spp_list, aes(x = area_ratio, y = sm_perc, color = sm_range)) + 
#   ggtheme_plot + 
#   geom_point() + 
#   geom_vline(x = area_ratio_quartiles,
#              color = 'grey50', linetype = 'dashed') +
#   geom_hline(y = pct_overlap_quartiles,
#              color = 'grey50', linetype = 'dashed') +
#   geom_point(data = spp_list_label, color = 'grey30') + 
#   geom_text(data = spp_list_label, 
#             aes(label = sciname_label), 
#             color = 'grey30',
#             fontface = 'bold.italic',
#             size = 2,
#             hjust = .25,
#             vjust = 1.5) +
#   labs(x = 'Range area ratio (smaller:larger) * 100%', 
#        y = 'Percent overlap smaller range in larger', 
#        title = 'Ratio of species range vs. percent overlap of \nsmaller range with larger')
# 
# scatter_arearatio_vs_overlap2 <- scatter_arearatio_vs_overlap2 + 
#   geom_text(x = 40, y = 5, label = lm_eqn2(spp_list), parse = TRUE, color = 'grey30')
# 
# ggsave('figures/AreaRatiovsPercOverlap2.png', 
#        width = 6, height = 4)
# 
# print(scatter_arearatio_vs_overlap2)

# summary(lm(area_ratio ~ sm_perc, spp_list))

@@

barchart_spp_gp_quads <- ggplot(spp_gp_quadrants %>% 
                                  transform(spp_group = reorder(spp_group, pct_q1)), 
                                aes(x = spp_group, fill = quad, weight = pct_quad)) +
  ggtheme_plot + 
  geom_bar(stat = 'bin', alpha = .5) +
  scale_fill_manual(values = c('q1' = 'green4', 'q2' = 'greenyellow', 'q3' = 'orange', 'q4' = 'orangered2')) +
  geom_text(aes(label = sprintf('n = %s', n_spp), y = .05), hjust = 0) +
  coord_flip() +
  labs(x = 'Species Group', 
       y = 'Relative number of species per quadrant', 
       title = 'Species group breakdown by quadrant')

@@

@@

Percent coverage as function of area, clipping to area below 100,000,000 km^2^:

``` {r, echo = FALSE, warning = FALSE, message = FALSE}
# remove all area greater than 100000000

@@

### create dataframe that gives me all species, their total number of cells, 
### total number of cells >= 0.4 prob, difference between the two and threat
spp_df3 <- spp_cells %>%
  group_by(species_id) %>%
  summarize(cells_0.4 = sum(probability >= 0.4),
            cells_t = sum(probability >= 0)) %>%
  left_join(spp, by = c('species_id' = 'SPECIESID')) %>%
  mutate(prop = cells_0.4 / cells_t,
         iucn_code = ifelse(is.na(iucn_code), 'NE', iucn_code)) %>%
  filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX', 'DD', 'NE')) %>%
  # filter(toupper(iucn_code) %in% c('LC', 'NT', 'VU', 'EN', 'CR', 'EX')) %>%
  select(sciname, species_id, cells_0.4, cells_t, prop, iucn_code)

iucn_levels = c('LC', 'NT', 'VU', 'EN', 'CR', 'EX', 'DD', 'NE')
spp_df3 <- spp_df3 %>% 
  transform(iucn_code = factor(iucn_code, levels = iucn_levels))

hist_area_vs_threshold <- ggplot(spp_df3, aes(prop, fill = iucn_code)) + 
  ggtheme_plot + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.05, alpha = .7) + 
  scale_fill_manual(values = c('LC' = 'green4', 
                               'NT' = 'greenyellow', 
                               'VU' = 'orange', 
                               'EN' = 'orangered2', 
                               'CR' = 'red4',
                               'DD' = 'grey60',
                               'NE' = 'grey70')) +
  labs(x = 'Proportion of range area included when selecting a threshold of 0.4', 
       y = sprintf('Proportion of all aquamaps species (n = %s)', nrow(spp_df3)))

@@

spp_df3a <- spp_df3 %>% 
  filter(!iucn_code %in% c('DD', 'NE'))

hist_area_vs_threshold2 <- ggplot(spp_df3a, aes(prop, fill = iucn_code)) + 
  ggtheme_plot + 
  geom_histogram(aes(y = ..count../sum(..count..)), binwidth = 0.05, alpha = .7) + 
  scale_fill_manual(values = c('LC' = 'green4', 
                               'NT' = 'greenyellow', 
                               'VU' = 'orange', 
                               'EN' = 'orangered2', 
                               'CR' = 'red4',
                               'DD' = 'grey60',
                               'NE' = 'grey70')) +
  labs(x = 'Proportion of range area included when selecting a threshold of 0.4', 
       y = sprintf('Proportion of all aquamaps species (n = %s)', nrow(spp_df3a)))

ggsave('figures/AMThreshold2.png', width = 6, height = 4)

print(hist_area_vs_threshold2)
